Efficient evaluation of threshold queries of derived fields in a numerical simulation database

### 2014-10-07 paper.tex
Bigger fonts for figures. Other minor changes.

\begin{abstract}
%Data-intensive architectures and compute clusters have been deployed to store, manage and provide public access to such datasets. 
%The systems built on top of these architectures, sometimes called virtual laboratories, usually do not support the analysis functionality that scientists are 
%interested in out of the box. Moreover, d
In this paper we present a method for the efficient evaluation of threshold queries of derived fields for large numerical 
simulation datasets stored in a cluster of relational databases. 
Such simulations are commonplace in the study of many different physical phenomena. 
The datasets produced by these simulations are in the TB and even PB ranges. 
Data-intensive computations that examine entire time-steps of the simulation data are impractical to perform
locally and can take days or months to iterate over the entire dataset. 
The integrated method for the evaluation of threshold queries that we have developed achieves scalability through data-parallel execution
of the computations on the nodes of the database cluster. 
We extend the database system with the introduction of an application-aware cache for query results, building on the concept of semantic caching.
The cache has little overhead and improves query performance by over an order of magnitude for queries that hit the cache.
\end{abstract}

@@

Database systems support rollup queries, including top-$k$ queries, but in most cases these
queries apply only simple linear score functions on the attribute values of individual records \cite{Ilyas}.
Additionally, many top-$k$ query evaluation techniques rely on the score functions being monotone in order to perform early pruning (see \cite{Ilyas} for a
survey of top-$k$ evaluation strategies).
This is an assumption that we cannot make for the functions used to compute all the different possible derived fields of interest in fluid mechanics.
Even approaches that aim to work with general score functions \cite{Deshpande, Xin} assume that the function operates on the attributes of a single record.
In contrast our approach computes the value of a derived field at each location in the array by examining the data at all neighboring locations, which
are within the kernel of the computation. 
Finally, a top-$k$ approach may not be suitable in the cases where scientists are interested in performing threshold queries at different time-steps as the
same threshold level will produce different number of points in the result set for different time-steps.

@@

%What are the key components of my approach and results? Also include any specific limitations.
We present a method for the efficient evaluation of threshold queries over fields derived from the raw vector or scalar fields of the numerical simulation 
stored in the database. The derived fields, which our method computes and operates on are derived by performing kernel computations over the array data 
stored in the database.
Our method makes effective use of the cluster resources and achieves high throughput and scalability. 
We exploit the parallelism 
available in the cluster by means of data-parallel execution of the computations. Each computation is distributed across the nodes of the cluster and
is executed in parallel by several processes within a database node by creating multiple connections to each node. We extend the database management
system with an
application-aware cache for query results. We build on the idea of an application-aware cache introduced by Lopez et al. \cite{Lopez} and more broadly 
on the concept of semantic caching \cite{Ren}. Rather than caching just data as is the case in system caches and the tree-cache described by Lopez et al.,
we cache the actual results of the computation, which leads to query performance improvement of over an order of magnitude.
The cache that we describe is therefore a limited version of a semantic cache.

@@

@@

\begin{table}[h]
\centering
\begin{tabular}{|c|c|c|c|c|} \hline
\multirow{2}{0.5in}{Threshold} & \multirow{2}{0.5in}{Points above threshold} & \multicolumn{3}{|c|}{Average Running time (s.)}\\ \cline{3-5}
& & No cache & \parbox[t]{0.55in}{With cache (miss)} & \parbox[t]{0.55in}{With cache (hit)} \\ \hline
80.0 & 4247 & 97.1 & 100.2 & 0.5\\ \hline
60.0 & 86580 & 113.7 & 115.9 & 1.2\\ \hline
44.0 & 909274 & 111.6 & 115.0 & 9.1\\ \hline
\end{tabular}
\caption{Effectiveness of caching.}
\label{tab:cache_effectiveness}
\end{table}

@@

Zhao et al. propose an algorithm for the processing of top-$k$ queries in large-scale distributed environments called BRANCA \cite{Zhao}. They build on the 
idea of semantic caching \cite{Ren} and make use of branch caches, which store results of previous top-$k$ queries with respect to the data stored on each server.
The caching mecahnism that we use is similar in that regard, but the queries that are evaluated in our system operate on derived fields, which are computed
at each location by accessing data from a surrounding region. The queries described in \cite{Zhao} operate over the attributes of individual records only using
simple linear score functions.

>>>>

\begin{abstract}
%Data-intensive architectures and compute clusters have been deployed to store, manage and provide public access to such datasets. 
%The systems built on top of these architectures, sometimes called virtual laboratories, usually do not support the analysis functionality that scientists are 
%interested in out of the box. Moreover, d
In this paper we present a method for the efficient evaluation of threshold queries of derived fields for large numerical 
simulation datasets stored in a cluster of relational databases. 
Such simulations are commonplace in the study of many different physical phenomena. 
The datasets produced by these simulations are in the TB and even PB ranges. 
Data-intensive computations that examine entire time-steps of the simulation data are impractical to perform
locally and can take days or months to iterate over the entire dataset. 
The integrated method for the evaluation of threshold queries that we have developed achieves scalability through data-parallel execution
of the computations on the nodes of the database cluster. 
We extend the scientific database cluster with the introduction of an application-aware cache for query results, building on the concept of semantic caching.
The cache has little overhead and improves query performance by over an order of magnitude for queries that hit the cache.
\end{abstract}

@@

Database systems support rollup queries, including top-$k$ queries, but in most cases these
queries apply only simple linear score functions on the attribute values of individual records.
Additionally, many top-$k$ query evaluation techniques rely on the score functions being monotone in order to perform early pruning (see \cite{Ilyas} for a
survey of top-$k$ evaluation strategies).
This is an assumption that we cannot make for the functions used to compute all the different possible derived fields of interest in fluid mechanics.
Even approaches that aim to work with general score functions \cite{Deshpande, Xin} assume that the function operates on the attributes of a single record.
In contrast our approach computes the value of a \emph{derived} field at each location in the array by examining the data at all neighboring locations, which
are within the kernel of the computation. The functions used to the derive the field may even be non-linear.
Finally, a top-$k$ approach may not be suitable in the cases where scientists are interested in performing threshold queries at different time-steps as the
same threshold level will produce different number of points in the result set for different time-steps.

@@

%What are the key components of my approach and results? Also include any specific limitations.
We present a method for the efficient evaluation of threshold queries over fields derived from the raw vector or scalar fields of the numerical simulation 
stored in the database. The derived fields, which our method computes and operates on are derived by performing kernel computations over the 
stored array data.
Our method makes effective use of the cluster resources and achieves high throughput and scalability. 
We exploit the parallelism 
available in the cluster by means of data-parallel execution of the computations. Each computation is distributed across the nodes of the cluster and
is executed in parallel by several processes within a database node by creating multiple connections to each node. We extend the database management
system with an
application-aware cache for query results. We build on the idea of an application-aware cache introduced by Lopez et al. \cite{Lopez} and more broadly 
on the concept of semantic caching \cite{Ren}. Rather than caching just data as is the case in system caches and the tree-cache described by Lopez et al.,
we cache the actual results of the computation, which leads to query performance improvement of over an order of magnitude.
The cache that we describe is therefore a limited version of a semantic cache.

@@

Not all query results are suitable to caching. Most of the queries submitted to the JHTDB other than threshold queries request data at a collection of target
locations. Given that there are $1024^4$ possible locations for three of the datasets and $6*1024^4$ locations for the channel-flow dataset the chance of
reuse for the results of these queries is extremely small. This is why the cache currently stores only the results of threshold queries and is therefore a 
limited version of a semantic cache. Nevertheless, it can easily be extended to cache the results of other query types as well if that becomes advantageous.

@@

\begin{table}[h]
\centering
\begin{tabular}{|c|c|c|c|c|} \hline
\multirow{2}{0.5in}{Vorticity threshold} & \multirow{2}{0.5in}{Points above threshold} & \multicolumn{3}{|c|}{Average Running time (s.)}\\ \cline{3-5}
& & No cache & \parbox[t]{0.55in}{With cache (miss)} & \parbox[t]{0.55in}{With cache (hit)} \\ \hline
80.0 & 4247 & 97.1 & 100.2 & 0.5\\ \hline
60.0 & 86580 & 113.7 & 115.9 & 1.2\\ \hline
44.0 & 909274 & 111.6 & 115.0 & 9.1\\ \hline
\end{tabular}
\caption{Effectiveness of caching.}
\label{tab:cache_effectiveness}
\end{table}

@@

Zhao et al. propose an algorithm for the processing of top-$k$ queries in large-scale distributed environments called BRANCA \cite{Zhao}. They build on the 
idea of semantic caching \cite{Ren} and make use of branch caches, which store results of previous top-$k$ queries with respect to the data stored on each server.
The caching mechanism that we use is similar in that regard, but the queries that are evaluated in our system operate on derived fields, which are computed
at each location by accessing data from a surrounding region. The queries described in \cite{Zhao} operate over the attributes of individual records only using
simple linear score functions.
